Simple and robust inference of connectomes using partial correlation coefficients

### 2014-06-10 article.tex
merge both version of filtering

\section{Signal filtering} \label{sec:filter}
% Explain why we need to filter
% TODO

\textcolor{red}{AS: Après avoir discuté avec Gilles, une autre façon de
présenter les choses qui est moins rigoureuse que l'approche de AJ mais plus
"motivée" --> Certains aspects pourraient être incorporés dans la version de AJ.}

Prior inferring the network, the fluorescent calcium time series
are smoothed using several linear and non linear filters.

% Information in burst not in value --> filtering with hard-threshold
% Considering the average delay between two neurons, low frequency event can be neglected --> high-pass filter
% Condidering noise, high frequency event can be neglected --> low-pass filter

%\paragraph{Motivation\\} 
Intuitively, connectivity information between two
neurons may be retrieved from successive and slightly delayed burst events. In
case of a directed relationship, the burst order should be the same while in
case of an undirected relationship, it might change. Therefore, the
fluorescent signal can be reduced to a binary signal representing the activity
of a neuron. This is the idea behind the filter $g_1$:  \begin{align}
g_1^q(X^i_t) = (X^i_t)^{0.9} \mathbb{1}(X^i_t \ge q) \end{align} where ... .

Considering the short delay for a communication between neurons, slow events
can be neglected to retrieve direct relationships. Note that the average time
delay is about $1ms$ or $2ms$ while the fluorescent signal time resolution is
$20ms$. Let us also highlight that even if two neurons are spatially far from
each other, the information is expected in the same time step or at most in
the next time step. It is therefore not necessary to keep low-frequency event
and we thus choose to consider the discrete derivative of the signal, i.e. a
high-pass filtered version, $g_2$:  \begin{align} g_2(X_t^i) = X^i_t -
X^i_{t-1} \end{align} where ... .

However, the signal, by the imaging capture process we suppose, is very noisy.
A low-pass filter seems a good way to smooth the signal and only keep relevant
informations. We thus apply the filter
$g_3$: 
\begin{align} g_3^k(X^i_t) =
\sum_{l=-3}^{2} c^k_l X^i_{t+l}  \end{align} where ... .
% We then define our preprocessing function $G$ as
% \begin{align}
% G(X_t^i) = g^q_1(g_3(g_2(X_t^i))).
% \end{align}

Then we apply a non-linear filter based on the current activity of the systems
in order to favor the association measure when a medium size set of neurons
are firing together, while artificially reducing the association measure when
the activity is extremely high, i.e. in case of a network burst. \begin{align}
h_1(X_t^i) = (X_t^i + 1)^{\left( 1 + \dfrac{1}{a_t} \right)} &\text{ with }
a_t = \sum_{i=1}^p \left ( X_t^i + \dfrac{1}{2} X_{t-1}^i \right ) \\
h_2(X_t^i) = (X_t^i)^{r(a_t)} &\text{ with } r_t = \frac{a_t}{\max_{l \in
T}{a_l}}  \end{align}

\textcolor{red}{miss the case where activity < 0->1} where $r(x)$ is a 4-piecewise continuous functions whose parameters were tuned
for each low-pass filter.

% Hence, we define our preprocessing function $H$ as
% \begin{align}
% H(X_t^i) = h_2 \bullet h_1 \bullet G(X).
% \end{align}

Hence, we define our preprocessing function $G$ as
\begin{align}
G^{q,k}(X_t^i) = h_2 \bullet h_1 \bullet g^q_1 \bullet g_2 \bullet g_3^k(X_t^i).
\end{align}

Finally since there is a high correlation between neurons, we made a
low-rank hypothesis and perform noise filtering through a principal component
approach [TODO cite]. From the $p$ eigen vectors, we only kept $80\%$ of
the eigen vectors which have the largest eigen value \textcolor{red}{fig?}.

% Filtering methodology 
Prior inferring the network, the fluorescent calcium
time series are smoothed using several linear and non linear filters. A low-
pass filter, a high-pass filter and a non linear hard-thresholding based
filter are applied for each possible time step $t$ % TODO fix first equation
is wrong due notation change \begin{align} X^\prime_{t,i} &= \sum_{l=-3}^2
c_l^k X_{t+l,i} \forall i, \\ X^{\prime\prime}_{t,i} &= X^{\prime}_{t,i} -
X^{\prime}_{t-1,i} \forall i, \\ X^{\prime\prime\prime}_{t,i} &=
{X^{\prime\prime}_{t,i}}^{0.9} I(X^{\prime\prime}_{t,i} \geq f_3) \forall i,
\end{align} where the $c_l^k$ coefficients were taken in one of the four
following tuples $\left\{c_l^1\right\}=(0, 0, 1, 1, 1, 1)$,
$\left\{c_l^2\right\}=(0, 0, 1, 1, 1, 0)$, $\left\{c_l^3\right\}=(1, 1, 1, 1,
0, 0)$ or $\left\{c_l^4\right\}=(0.4, 0.8, 1, 1, 0, 0)$ and $f_3$ is a
parameter of the hard-threshold filter.

Then we apply on all remaining timesteps $T$ a non-linear filter based on the current
activity of the systems in order to favor the association measure when
a medium size set of neurons are firing together, while artificially
reducing the association measure in case of extreme network activity,
i.e. high when a network firing is happening and low when no neuron is
active:
\begin{align}
a_t &= \sum_{i=1}^p X^{\prime\prime\prime}_{t,i} +
                   \frac{1}{2} X^{\prime\prime\prime}_{t-1,i} \forall t \in T\\
r_t &= \frac{a_t}{\max_{l \in T}{a_l}} \\
X^{\prime\prime\prime\prime}_{t,i} &=
\left\{
  \begin{array}{l}
    1  : r_t < 0\\
    {(X^{\prime\prime\prime}_{t,i} + 1)^{(a_t^{-1} + 1)}}^{g(r_t)} : r_t \geq 0
  \end{array}
\right.
\end{align}
where $g(x) = b_1 I(f_4 < x < f_5) + b_2  I(f_5 \leq x < f_6) +
b_3 (I(x\leq f_4) + I(f_ 6 \leq x))$ is a piecewise continuous function of
parameters $b_1, b_2, b_3, f_4, f_5, f_6$. The coefficient were tuned
for each set of coefficient $\left\{c_l^k\right\}$.

Finally since there is a high correlation between neurons, we made a
low-rank hypothesis and perform noise filtering through a principal component
approach [TODO cite]. From the $p$ eigen vectors, we only kept $80\%$ of
the eigen vectors which have the largest eigen value.

>>>>

\section{Signal filtering} \label{sec:filter}

% Explain why we need to filter

Prior inferring the network, the fluorescent calcium time series
are smoothed using several linear and non linear filters.
Considering the short delay for a communication between neurons, slow events
can be neglected to retrieve direct relationships. The average time
delay is indeed about $1ms$ or $2ms$ while the fluorescent signal time
resolution is $20ms$. Let us also highlight that even if two neurons are
spatially far from each other, the information is expected in the same time
step or at most in the next time step. Thus very low frequency event can be neglected.
However, the image fluorescence calcium signal is also very noisy. Thus,
A low-pass filter (Equation \ref{eq:low-pass}) is first applied followed by a
high-pass filter (Equation \ref{eq:high-pass}), the discrete derivative of
the signal,
\begin{align}
X^\prime_{t,i} &= \sum_{l=-3}^2 c_l^k X_{t+l,i} \forall i, \label{eq:low-pass}\\
X^{\prime\prime}_{t,i} &= X^{\prime}_{t,i} - X^{\prime}_{t-1,i} \forall i, \label{eq:high-pass}
\end{align}
where the $c_l^k$ coefficients were taken in one of the four
following tuples
$\left(c_l^1\right)_{l=-3}^2=(0, 0, 1, 1, 1, 1)$,
$\left(c_l^2\right)_{l=-3}^2=(0, 0, 1, 1, 1, 0)$,
$\left(c_l^3\right)_{l=-3}^2=(1, 1, 1, 1, 0, 0)$ or
$\left(c_l^4\right)_{l=-3}^2=(0.4, 0.8, 1, 1, 0, 0)$.

Connectivity information between two neurons may be retrieved from successive
and slightly delayed burst events. At a local level, direct link
between two neurons will have shorter delay between fluorescence
peaks than undirected links. However, the fluorescence signal has a slow
decay compare to its rise. Therefore, we filter those unrelevant information
with a hard thresholding filter
\begin{align}
X^{\prime\prime\prime}_{t,i} &=
{X^{\prime\prime}_{t,i}}^{0.9} I(X^{\prime\prime}_{t,i} \geq f_3) \forall i
\label{eq:hard-treshold-filter}
\end{align}
where $f_3$ is a parameter of the hard-threshold filter and the $0.9$
exponent reduce the difference between peaks values.

Then we apply on all remaining timesteps $T$ a non-linear filter based on the current
activity of the systems in order to favor the association measure when
a medium size set of neurons are firing together, while artificially
reducing the association measure in case of extreme network activity,
i.e. high when a network firing is happening and low when no neuron is
active:
\begin{equation}
X^{\prime\prime\prime\prime}_{t,i} =
\left\{
  \begin{array}{l}
    1  : r_t < 0\\
    {(X^{\prime\prime\prime}_{t,i} + 1)^{(a_t^{-1} + 1)}}^{g(r_t)} : r_t \geq 0
  \end{array}
\right.
\text{ with }
\left\{
  \begin{array}{ll}
    a_t &= \sum_{i=1}^p X^{\prime\prime\prime}_{t,i} +
                       \frac{1}{2} X^{\prime\prime\prime}_{t-1,i} \forall t \in T\\
    r_t &= \frac{a_t}{\max_{l \in T}{a_l}} \\
  \end{array}
\right.
\end{equation}
where $g(x) = b_1 I(f_4 < x < f_5) + b_2  I(f_5 \leq x < f_6) +
b_3 (I(x\leq f_4) + I(f_ 6 \leq x))$ is a piecewise continuous function of
parameters $b_1, b_2, b_3, f_4, f_5, f_6$. The coefficient were tuned
for each set $i\in\left\{1, 2, 3, 4\right\}$ of coefficient
$\left(c_l^i\right)_{l=-3}^2$.

Finally since there is a high correlation between neurons, we made a
low-rank hypothesis and perform noise filtering through a principal component
approach [TODO cite]. From the $p$ eigen vectors, we only kept $80\%$ of
the eigen vectors which have the largest eigen value. In the context of the
challenge, this assumption was verified on all provided dataset
